Shift-sensitive Power Transformations for Transforming Data to Normality

### 2023-11-07 manuscript/manuscript.Rmd
Added plot that shows effect of outliers.
Fixed labels.

```{r decreased-normality, echo=FALSE, fig.cap=cap, fig.height=2.0, warning=FALSE}
cap <- paste0(
  "\\label{fig:decreased-normality}",
  "Power transformations may reduce normality depending on location. ",
  "Variable $x$ consists of $10000$ values randomly drawn from a normal distribution $\\mathcal{N}(\\mu, 1)$. ",
  "For both Box-Cox and Yeo-Johnson transformations, we would expect $\\lambda = 1$, since $x$ is already normally distributed. ",
  "The figure shows that instead, $\\lambda$ is strongly dependent on the location $\\mu$, ",
  "and even breaks down due to numerical issues for $\\mu > 5 \\cdot 10^4$.")

@@

```{r sensitivity-to-outliers, echo=FALSE, fig.cap=cap, fig.height=2.0, warning=FALSE}
cap <- paste0(
  "\\label{fig:sensitivity-to-outliers}",
  "Power transformations are affected by outliers. ",
  "Variable $x$ consists of $1000$ values randomly drawn from a normal distribution $\\mathcal{N}(\\mu, 1)$, ",
  "with a single outlier with value $d$. ",
  "For both Box-Cox and Yeo-Johnson transformations, we would expect $\\lambda = 1$, since $x$ is already normally distributed. ",
  "The figure shows that instead, $\\lambda$ is strongly affected by the presence of the outlier at $d$, ",
  "and even breaks down due to numerical issues for $d > 5 \\cdot 10^4$.")

@@

```{r weighting-functions, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=2.5}
cap <- paste0(
  "\\label{fig:weighting-functions}",
  "Weighting functions investigated in this study to make power transformations more robust against outliers. ",
  "The step function was parameterised with $k_1 = 0.60$. ",
  "The triangle and tapered cosine functions were both parameterised with $k_1 = 0.30$ and $k_2 = 0.90$.")

@@

Realistic feature distributions may be skewed. Gijbels et al. describe a recipe for introducing skewness into the otherwise symmetric generalised norm distribution [@Gijbels2019-te], leading to the following probability density function:

@@

To assess whether the modified power transformations lead to $\lambda$ values that are invariant to shifts, we simulate three different distributions. The first distribution is a normal distribution, as previously shown in Figure \ref{fig:decreased-normality}.
We first randomly draw $10000$ values from a normal distribution: $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)$, or equivalently $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)$.
The second distribution is a right-skewed normal distribution $\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)$.
The third distribution is a left-skewed normal distribution $\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)$.
We then compute transformation parameter $\hat{\lambda}$ using the original definitions (equations \ref{box-cox-original} and \ref{yeo-johnson-original}) and the shift-sensitive definitions (equations \ref{box-cox-shift} and \ref{yeo-johnson-shift}) for each distribution, after adding a positive value $d$ to each distribution. Here, $d \in [1, 10^6]$.

```{r shifted-distributions, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "\\label{fig:shifted-distributions}",
  "Shift-sensitive power transformation produces transformation parameters that are invariant to location. ",
  "Samples were drawn from normal, right-skewed and left-skewed distributions, respectively,  which then underwent a shift $d$. ",
  "Estimates of transformation parameter $\\lambda$ with the original power transformations (circle) show strong dependency on the overall location of the distribution, ",
  "whereas estimates with the shift-sensitive power transformations (triangle) are constant.")

@@

```{=latex}
\begin{table}
\begin{center}
\label{tab:optimal-weighting-parameters-box-cox}
\caption{Optimal weighting parameters and corresponding loss for \textbf{shift-sensitive Box-Cox} power transformations. $p^{*}$ indicates use of the empirical distribution of feature values, $z$ the z-score of the transformed feature values, and $r$ the residual error between the z-score of transformed feature values and the expected z-score according to the normal distribution. The \textit{initial} column shows the starting parameter value for the optimisation process, with the corresponding boundary values in the \textit{limits} column. The {optimal} column shows the optimal parameter values. The \textit{loss} column shows the loss achieved by each method, under optimised parameters. Lower loss indicates better robustness against outliers.}
\begin{tabular}{l r r r r r r r}

@@

```{=latex}
\begin{table}
\begin{center}
\label{tab:optimal-weighting-parameters-yeo-johnson}
\caption{Optimal weighting parameters and corresponding loss for \textbf{shift-sensitive Yeo-Johnson} power transformations. $p^{*}$ indicates use of the empirical distribution of feature values, $z$ the z-score of the transformed feature values, and $r$ the residual error between the z-score of transformed feature values and the expected z-score according to the normal distribution. The \textit{initial} column shows the starting parameter value for the optimisation process, with the corresponding boundary values in the \textit{limits} column. The {optimal} column shows the optimal parameter values. The \textit{loss} column shows the loss achieved by each method, under optimised parameters. Lower loss indicates better robustness against outliers.}
\begin{tabular}{l r r r r r r r}

@@

```{r optimised-weighting-function-parameters, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "\\label{fig:optimised-weighting-function-parameters}",
  "Robustness of power transformations after optimising weighting function parameters. ",
  "The distribution of errors, i.e. the difference between robustly fitted $\\hat{\\lambda}^r$ ",
  "and expected $\\hat{\\lambda}_{0}$ found in the absence of outliers, is shown for 1000 randomly ",
  "parametrised asymmetric generalised normal distributions with up to 10% outliers on either side ",
  "of the distribution. The top and bottom panel show errors for the shift-sensitive Box-Cox transformation, ",
  "and shift-sensitive Yeo-Johnson transformation, respectively. ",
  "In each panel, the error distribution for the non-robust transformation is shown to the left ",
  "for comparison with different weighting methods. ",
  "Note that a square root transformation was used along the $y$-axis. Median errors are shown above each distribution."
)

@@

Figure \ref{tab:empirical-central-normality-residual-error} shows the residual errors of the transformed distribution as function of the empirical probability. As expected, the largest deviations from normality appear on the extremities of this range. For Box-Cox, the effect of outliers on the lower end of the distribution leads to noticeable asymmetry. Between very low and very high empirical probabilities, residual errors are constrained and relatively flat. This is the candidate range for the central portion of the data.

```{r empirical-central-normality-residual-error, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "\\label{tab:empirical-central-normality-residual-error}",
  "Residual error as a function of empirical probability. ",
  "Percentiles of the error are shown for robust, shift-sensitive transformations of 10000 randomly drawn asymmetric ",
  "generalised normal distributions, with 10 randomly drawn outliers distributions, ",
  "(up to 10% of samples) each. Larger errors occur for more extreme values at the edges of each distribution. ",
  "For comparison, the 99th percentile of residual errors for transformations of outlier-free distributions are shown (grey dashed line).")

@@

```
We then consider the empirical probability of type I errors: the probability of incorrectly classifying a centrally normal distribution as being non-normal based on the test statistic. Under the assumption that the asymmetric generalised normal distributions are centrally normal after robust transformation, this produces the relationship shown in Figure \ref{tab:empirical-central-normality-type-1-error-rate}. Since for $\kappa \leq 0.80$ error curves are similar, we fix the value of the central portion $\kappa$ to 0.80. This leads to test statistic values listed in Table \ref{tab:empirical-central-normality}, which derive from the type I error rates of the Yeo-Johnson transformation.

```{r empirical-central-normality-type-1-error-rate, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "\\label{tab:empirical-central-normality-type-1-error-rate}",
  "Type 1 error rate of transformed asymmetric generalised normal distributions as function of ",
  "the test statistic for the central portion $\\kappa$ of the distribution.")

@@

**MISSING TABLE DATA FOR YEO-JOHNSON**

```{=latex}
\begin{table}
\begin{center}
\label{tab:empirical-central-normality}
\caption{Probability of central normality, based on type I error rates for Box-Cox transformations.}
\begin{tabular}{c c c c c c c}

\toprule
0.50 & 0.20 & 0.10 & 0.05 & 0.02 & 0.01 & 0.001 \\

\midrule
0.042 & 0.063 & 0.077 & 0.090 & 0.106 & 0.118 & 0.159 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}
```

@@

```{r empirical-central-normality-examples, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "\\label{fig:empirical-central-normality-examples}",
  "Bi-modal distributions and empirical goodness of fit test results. ",
  "The feature (black) is a mixture of two identical sample sets (blue and orange) ",
  "drawn from normal distributions that are offset by a distance $d$. ",
  "We use the empirical goodness of fit test to compute p-values for the hypothesis that ",
  "the distribution is centrally normal. As may be observed, with increasing offset $d$ ",
  "the probability that the feature is centrally normal decreases. ",
  "Quantile-quantile plots are drawn below each distribution."
)

@@

```{r top-gear, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=2}
cap <- paste0(
  "\\label{fig:top-gear}",
  "Quantile-quantile plots for the fuel consumption data in the Top Gear dataset. ",
  "This data contains three strong outliers for highly efficient vehicles. ",
  "The left panel shows the quantile-quantile plot for the original data. ",
  "The middle panel shows the quantile-quantile plots for data transformed using ",
  "the original Yeo-Johnson transformation, and Raymaekers and Rousseeuw's robust adaptation [@Raymaekers2021-kq]. ",
  "For these, the empirical central normality test yields $p=0.01$, and $p=0.58$, respectively. ",
  "The right panels shows the quantile-quantile plots for data transformed using ",
  "the non-robust and robust shift-sensitive transformations. ",
  "For these, the empirical central normality test yields $p=0.01$, and $p=0.36$, respectively. ",
  "Samples with observed quantiles below $-3.0$ or above $3.0$ are indicated by crosses."
)

>>>>

```{r decreased-normality, echo=FALSE, fig.cap=cap, fig.height=2.0, warning=FALSE}
cap <- paste0(
  "Power transformations may reduce normality depending on location. ",
  "Variable $x$ consists of $10000$ values randomly drawn from a normal distribution $\\mathcal{N}(\\mu, 1)$. ",
  "For both Box-Cox and Yeo-Johnson transformations, we would expect $\\lambda = 1$, since $x$ is already normally distributed. ",
  "The figure shows that instead, $\\lambda$ is strongly dependent on the location $\\mu$, ",
  "and even breaks down due to numerical issues for $\\mu > 5 \\cdot 10^4$.")

@@

```{r sensitivity-to-outliers, echo=FALSE, fig.cap=cap, fig.height=2.0, warning=FALSE}
cap <- paste0(
  "Power transformations are affected by outliers. ",
  "Variable $x$ consists of $1000$ values randomly drawn from a normal distribution $\\mathcal{N}(\\mu, 1)$, ",
  "with a single outlier with value $d$. ",
  "For both Box-Cox and Yeo-Johnson transformations, we would expect $\\lambda = 1$, since $x$ is already normally distributed. ",
  "The figure shows that instead, $\\lambda$ is strongly affected by the presence of the outlier at $d$.")

@@

```{r weighting-functions, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=2.5}
cap <- paste0(
  "Weighting functions investigated in this study to make power transformations more robust against outliers. ",
  "The step function was parameterised with $k_1 = 0.60$. ",
  "The triangle and tapered cosine functions were both parameterised with $k_1 = 0.30$ and $k_2 = 0.90$.")

@@

Realistic feature distributions may be skewed. Gijbels et al. describe a recipe for introducing skewness into the otherwise symmetric generalised normal distribution [@Gijbels2019-te], leading to the following probability density function:

@@

To assess whether the modified power transformations lead to $\lambda$ values that are invariant to shifts, we simulate three different distributions. The first distribution is a normal distribution, as previously shown in Figure \ref{fig:decreased-normality}.
We first randomly draw $10000$ values from a normal distribution: $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)$, or equivalently $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)$.
The second distribution is a right-skewed normal distribution $\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)$.
The third distribution is a left-skewed normal distribution $\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)$.
We then compute transformation parameter $\hat{\lambda}$ using the original definitions (equations \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}) and the shift-sensitive definitions (equations \ref{eqn:box-cox-shift} and \ref{eqn:yeo-johnson-shift}) for each distribution, after adding a positive value $d$ to each distribution. Here, $d \in [1, 10^6]$.

```{r shifted-distributions, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Shift-sensitive power transformation produces transformation parameters that are invariant to location. ",
  "Samples were drawn from normal, right-skewed and left-skewed distributions, respectively,  which then underwent a shift $d$. ",
  "Estimates of transformation parameter $\\lambda$ with the original power transformations (circle) show strong dependency on the overall location of the distribution, ",
  "whereas estimates with the shift-sensitive power transformations (triangle) are constant.")

@@

```{=latex}
\begin{table}
\begin{center}
\caption{Optimal weighting parameters and corresponding loss for \textbf{shift-sensitive Box-Cox} power transformations. $p^{*}$ indicates use of the empirical distribution of feature values, $z$ the z-score of the transformed feature values, and $r$ the residual error between the z-score of transformed feature values and the expected z-score according to the normal distribution. The \textit{initial} column shows the starting parameter value for the optimisation process, with the corresponding boundary values in the \textit{limits} column. The {optimal} column shows the optimal parameter values. The \textit{loss} column shows the loss achieved by each method, under optimised parameters. Lower loss indicates better robustness against outliers.}
\label{tab:optimal-weighting-parameters-box-cox}
\begin{tabular}{l r r r r r r r}

@@

```{=latex}
\begin{table}
\begin{center}
\caption{Optimal weighting parameters and corresponding loss for \textbf{shift-sensitive Yeo-Johnson} power transformations. $p^{*}$ indicates use of the empirical distribution of feature values, $z$ the z-score of the transformed feature values, and $r$ the residual error between the z-score of transformed feature values and the expected z-score according to the normal distribution. The \textit{initial} column shows the starting parameter value for the optimisation process, with the corresponding boundary values in the \textit{limits} column. The {optimal} column shows the optimal parameter values. The \textit{loss} column shows the loss achieved by each method, under optimised parameters. Lower loss indicates better robustness against outliers.}
\label{tab:optimal-weighting-parameters-yeo-johnson}
\begin{tabular}{l r r r r r r r}

@@

```{r optimised-weighting-function-parameters, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Robustness of power transformations after optimising weighting function parameters. ",
  "The distribution of errors, i.e. the difference between robustly fitted $\\hat{\\lambda}^r$ ",
  "and expected $\\hat{\\lambda}_{0}$ found in the absence of outliers, is shown for 1000 randomly ",
  "parametrised asymmetric generalised normal distributions with up to 10% outliers on either side ",
  "of the distribution. The top and bottom panel show errors for the shift-sensitive Box-Cox transformation, ",
  "and shift-sensitive Yeo-Johnson transformation, respectively. ",
  "In each panel, the error distribution for the non-robust transformation is shown to the left ",
  "for comparison with different weighting methods. ",
  "Note that a square root transformation was used along the $y$-axis. Median errors are shown above each distribution."
)

@@

Figure \ref{fig:empirical-central-normality-residual-error} shows the residual errors of the transformed distribution as function of the empirical probability. As expected, the largest deviations from normality appear on the extremities of this range. For Box-Cox, the effect of outliers on the lower end of the distribution leads to noticeable asymmetry. Between very low and very high empirical probabilities, residual errors are constrained and relatively flat. This is the candidate range for the central portion of the data.

```{r empirical-central-normality-residual-error, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Residual error as a function of empirical probability. ",
  "Percentiles of the error are shown for robust, shift-sensitive transformations of 10000 randomly drawn asymmetric ",
  "generalised normal distributions, with 10 randomly drawn outliers distributions, ",
  "(up to 10% of samples) each. Larger errors occur for more extreme values at the edges of each distribution. ",
  "For comparison, the 99th percentile of residual errors for transformations of outlier-free distributions are shown (grey dashed line).")

@@

```
We then consider the empirical probability of type I errors: the probability of incorrectly classifying a centrally normal distribution as being non-normal based on the test statistic. Under the assumption that the asymmetric generalised normal distributions are centrally normal after robust transformation, this produces the relationship shown in Figure \ref{fig:empirical-central-normality-type-1-error-rate}. Since for $\kappa \leq 0.80$ error curves are similar, we fix the value of the central portion $\kappa$ to 0.80. This leads to test statistic values listed in Table \ref{tab:empirical-central-normality}, which derive from the type I error rates of the Yeo-Johnson transformation.

```{r empirical-central-normality-type-1-error-rate, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Type 1 error rate of transformed asymmetric generalised normal distributions as function of ",
  "the test statistic for the central portion $\\kappa$ of the distribution.")

@@

```{=latex}
\begin{table}
\begin{center}
\caption{Test statistic for empirical central normality at $\kappa = 0.80$ as a function of Type I error rate.}
\label{tab:empirical-central-normality}
\begin{tabular}{l c c c c c c c}

\toprule
transformation & 0.50 & 0.20 & 0.10 & 0.05 & 0.02 & 0.01 & 0.001 \\

\midrule
rob. shift-sensitive Box-Cox     & 0.046 & 0.072 & 0.090 & 0.106 & 0.126 & 0.140 & 0.189 \\
rob. shift-sensitive Yeo-Johnson & 0.042 & 0.063 & 0.077 & 0.090 & 0.106 & 0.118 & 0.159 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}
```

@@

```{r empirical-central-normality-examples, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Bi-modal distributions and empirical goodness of fit test results. ",
  "The feature (black) is a mixture of two identical sample sets (blue and orange) ",
  "drawn from normal distributions that are offset by a distance $d$. ",
  "We use the empirical goodness of fit test to compute p-values for the hypothesis that ",
  "the distribution is centrally normal. As may be observed, with increasing offset $d$ ",
  "the probability that the feature is centrally normal decreases. ",
  "Quantile-quantile plots are drawn below each distribution."
)

@@

```{r top-gear, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=2}
cap <- paste0(
  "Quantile-quantile plots for the fuel consumption data in the Top Gear dataset. ",
  "This data contains three strong outliers for highly efficient vehicles. ",
  "The left panel shows the quantile-quantile plot for the original data. ",
  "The middle panel shows the quantile-quantile plots for data transformed using ",
  "the original Yeo-Johnson transformation, and Raymaekers and Rousseeuw's robust adaptation [@Raymaekers2021-kq]. ",
  "For these, the empirical central normality test yields $p=0.01$, and $p=0.58$, respectively. ",
  "The right panels shows the quantile-quantile plots for data transformed using ",
  "the non-robust and robust shift-sensitive transformations. ",
  "For these, the empirical central normality test yields $p=0.01$, and $p=0.36$, respectively. ",
  "Samples with observed quantiles below $-3.0$ or above $3.0$ are indicated by crosses."
)
